---
title: "2348141_Assignment 4"
author: "Raghotham (2348141)"
date: "2024-10-22"
output:
  word_document: default
  html_document: default
---

## Introduction 

Spatial autocorrelation is a statistical measure that describes the degree of dependency between observations in a spatial dataset. It measures how similar or different nearby observations are. In this analysis, we will use the "elect80" dataset from the "spData" package to examine spatial autocorrelation in the context of electoral data.

By calculating Global and Local Moran's I, we can assess both overall and localized patterns of spatial autocorrelation. This will help us identify areas with high or low electoral outcomes that are clustered or dispersed, providing insights into the spatial distribution of electoral phenomena.

## Objective

Using a dataset “elect80” from the “spData” package perform following:

1. Visualize the spatial distribution of the chosen variable

2. Calculate Global Moran’s I to assess overall spatial autocorrelation and interpret the results.

3. Further, identify local clusters by calculating Local Moran’s I to detect areas of significant local spatial autocorrelation.

4. Identify the Hot spots and Cold spot areas.

5. Create and interpret a Moran’s scatter plot, explaining how it highlights high-high, low- low, and outlier regions.

6. Present your findings with appropriate maps and plots, interpreting both global and local spatial patterns.

## Data Description :

A data set for 1980 Presidential election results covering 3,107 US counties using geographical
coordinates. A SpatialPointsDataFrame with 3107 observations on the following 7 variables.

- **FIPS**: a factor of county FIPS codes

- **long**: a numeric vector of longitude values

- **lat**: a numeric vector of latitude values

- **pc_turnout**: Votes cast as proportion of population over age 19 eligible to vote

- **pc_college**: Population with college degrees as proportion of population over age 19 eligible to vote

- **pc_homeownership**: Homeownership as proportion of population over age 19 eligible to vote

- **pc_income**: Income per capita of population over age 19 eligible to vote

## Analysis

### 1. 

```{r}
# import necessary libraries
library(spdep)
library(sf)
library(sp)
library(spData)
library(tmap)
library(spatialreg)
```

```{r}
data(elect80)

# Extract coordinates from the SpatialPointsDataFrame
coords <- coordinates(elect80)

# Create a k-nearest neighbors list (e.g., 4 nearest neighbors)
knn_nb <- knearneigh(coords, k = 4)
neighbors <- knn2nb(knn_nb)

# Convert the neighbors list to a spatial weights matrix
listw <- nb2listw(neighbors)

tm_shape(elect80) +
  tm_dots("pc_turnout", palette = "-RdYlGn", title = "Voter Turnout in 1980") +
  tm_layout(title = "Spatial Distribution of Turnout", legend.outside = TRUE)
```

**Interpretation**

The spatial distribution map of voter turnout in 1980 shows how turnout values vary across different regions. This visualization helps identify initial spatial patterns, such as whether voter turnout is uniformly distributed or whether there are visible clusters.

### 2.

```{r}
# Global Moran's I calculation
global_morans_i <- moran.test(elect80$pc_turnout, listw)

print(global_morans_i)

```

**Interpertation**

- The Global Moran's I for Turnout is 0.62440318605, indicating a moderate positive spatial autocorrelation, meaning regions with similar turnout values (high or low) are mostly clustered together. 

- The expected Moran's I under the null hypothesis of no spatial autocorrelation is -0.0003, while the observed value is significantly higher. 

- The test statistic's standard deviate of 51.271 and the p-value of < 2.2e-16 show that this result is highly statistically significant, rejecting the null hypothesis. 

- Turnout is not randomly distributed but exhibits significant clustering across the regions.

### 3.
```{r}
# Local Moran's I calculation
local_morans_i <- localmoran(elect80$pc_turnout, listw)

# Add the results to the dataset
elect80$local_morans_i <- local_morans_i[, 1]
elect80$p_value <- local_morans_i[, 5]

# Set significance threshold for cluster types
elect80$cluster_type <- ifelse(elect80$p_value < 0.05,
                               ifelse(elect80$local_morans_i > 0, "Hot Spot", "Cold Spot"),
                               "Not Significant")

# Visualize Local Moran's I using tm_dots
tm_shape(elect80) +
  tm_dots(col = "local_morans_i", style = "quantile", 
          palette = "-RdBu", title = "Local Moran's I") +
  tm_layout(main.title = "Local Moran's I for Vote Percentage")

```

**Interpreutation**

We see the Local Moran's I for voter turnout and adds the results to the dataset, identifying local spatial autocorrelation. Significant positive values indicate Hot Spots (high voter turnout clusters), while negative values signify Cold Spots (low turnout clusters). Regions where the p-value is above 0.05 are classified as Not Significant.

### 4.

```{r}
# Classify regions as Hotspot, Coldspot, or Not significant
elect80$cluster <- factor(ifelse(local_morans_i[, 5] < 0.05 & local_morans_i[, 1] > 0, "Hotspot",
                                 ifelse(local_morans_i[, 5] > -0.05 & local_morans_i[, 1] < 0, "Coldspot", "Not significant")))

# Visualize hotspots and coldspots using tm_dots for point data
tm_shape(elect80) +
  tm_dots(col = "cluster", palette = c("Hotspot" = "red", "Coldspot" = "blue", "Not significant" = "grey")) +
  tm_layout(main.title = "Hotspots and Coldspots of Turnout")
```

**Interpertation**

- **Hotspots** - The regions that are indicated as red are known as hotspots which means the regions where both the area and its neighbors have high values for the variable, indicating spatial clustering of intense activity. In voter turnout, hotspots reflect areas with high participation and strong local engagement.

- **Coldspots** - The region that are indicated as blue dots are called as coldspot region. Coldspots are regions where both the area and its neighbors have low values, indicating local clustering of low activity. In voter turnout, coldspots highlight areas with low participation, possibly facing barriers to engagement.

### 5.

```{r}
# Moran's Scatter Plot
moran.plot(elect80$pc_turnout, listw, labels = elect80$FIPS, 
           xlab = "pc_turnout", ylab = "Spatial Lag",
           main = "Moran's Scatterplot")
```

**Interpertation**

The Moran’s scatterplot visually represents the relationship between Turnouts and its spatial lag. The plot is divided into four quadrants:

- **Top-right quadrant (High-High):** Regions with high Turnouts are surrounded by neighbors with similarly high values, indicating positive spatial autocorrelation.

- **Bottom-left quadrant (Low-Low):** Regions with low Turnouts are clustered near other low-turnout regions, reinforcing clustering.

- **Top-left (Low-High)** and **Bottom-right (High-Low):** These outliers indicate regions where Turnouts differs from their neighbors.

```{r}
quadrant <- vector("numeric", length = nrow(local_morans_i))
mean_lifeExp <- mean(elect80$pc_turnout)
# Quadrant classification based on local moran's I values and mean
quadrant[elect80$pc_turnout > mean_lifeExp & local_morans_i[, 1] > 0] <- "High-High"
quadrant[elect80$pc_turnout < mean_lifeExp & local_morans_i[, 1] > 0] <- "Low-Low"
quadrant[elect80$pc_turnout > mean_lifeExp & local_morans_i[, 1] < 0] <- "High-Low"
quadrant[elect80$pc_turnout < mean_lifeExp & local_morans_i[, 1] < 0] <- "Low-High"
# Add the quadrant classification to the dataset
elect80$quadrant <- factor(quadrant, levels = c("High-High", "Low-Low", "High-Low", "Low-High"))
# Map of Local Moran's I clusters
tm_shape(elect80) +
  tm_dots("quadrant", palette = c("darkgreen", "darkred", "blue", "pink"), title = "Local Spatial Clusters") +
  tm_layout(title = "Clusters Based on Local Moran's I", legend.outside = TRUE)
```

**Interpertation**

This world map depicts local spatial clusters using Moran's I statistic, revealing patterns of an unspecified variable. It shows clear regional trends: 

- High-High clusters (dark green) indicates areas with high values surrounded by similar neighbors.

- Low-Low clusters (red) are showing regions with low values adjacent to other low-value areas.

- High-Low outliers (pink) appear representing high-value areas surrounded by low-value neighbors.

- Low-High outliers (Blue) are scattered uniformly indicating high areas with low neighbouring region.

## Conclusion

In this analysis of spatial autocorrelation using the "elect80" dataset, we observed significant clustering patterns in voter turnout across US counties. The calculation of Global Moran’s I revealed a moderate positive spatial autocorrelation, indicating that regions with similar turnout values tend to be grouped together. This finding underscores the importance of understanding spatial dependencies in electoral data, as they can influence political engagement and outcomes.

Local Moran’s I analysis further identified specific clusters, categorizing regions into hotspots and coldspots. Hotspots, characterized by high turnout values surrounded by similar neighbors, reflect areas of strong civic engagement, while coldspots indicate regions with low participation. These localized insights are crucial for policymakers and electoral strategists aiming to enhance voter engagement and address barriers in areas with low turnout.

Overall, the visualization of spatial patterns through maps and scatter plots provided a comprehensive understanding of the electoral landscape. The findings highlight the necessity of considering spatial factors in electoral studies, as they reveal underlying trends that may not be apparent through traditional analysis. By leveraging spatial statistics, stakeholders can better target interventions and foster greater electoral participation across diverse communities.